\Large The state learner -- a super learner for right-censored data

### 2024-05-27 tex/journal-version/manuscript.tex
reformatting

%% Packages
\usepackage{amsthm, dsfont, amssymb,soul,xcolor,
  % amsfonts,
  enumitem,
  amsmath,graphicx,verbatim,pgfplots,tikz,prodint,booktabs}

@@

% % Handling new lines
% \setlength{\parskip}{1em}
% \setlength{\parindent}{0em}

% % Handling space after sections
% \usepackage{titlesec}
% \titlespacing*{\section}{0em}{2em}{0em}
% \titlespacing*{\subsection}{0em}{2em}{0em}
% \titlespacing*{\subsubsection}{0em}{2em}{0em}

% % No spacing after in start of list
% \setlist[itemize]{topsep=0pt}
% \setlist[enumerate]{topsep=0pt}

@@

\theoremstyle{thmstyleone}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
% \newtheorem{proposition}[theorem]{Proposition}%
% to get separate numbers for theorem and proposition etc.:
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

% \usepackage[tablesonly,heads,nolists]{endfloat}

%% change enumerate labels
% \renewcommand{\labelenumi}{\normalfont (\alph{enumi})}

%% Font size etc
\linespread{1.65}

\usepackage{lineno}
\linenumbers

@@

\RequirePackage{graphicx}
% \RequirePackage{caption}
% \RequirePackage{array}
% \RequirePackage{color}
% \RequirePackage{xcolor}
% \RequirePackage{amssymb}
% \RequirePackage{flushend}
% \RequirePackage{stfloats}
% \RequirePackage[figuresright]{rotating}
% \RequirePackage{chngpage}
% \RequirePackage{totcount}
% \RequirePackage{fix-cm}
\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

% New operators and commands
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
% \newcommand{\G}[1][]{%
%    \ifthenelse{ \equal{#1}{} }
%       {\ensuremath{\mathbb{G}_n}}
%       {\ensuremath{\mathbb{G}_{#1}}}
% }
% New version:
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure

\newcommand{\data}{\ensuremath{\mathcal{D}}}

%% Title
\title{The state learner -- a super learner for right-censored data}

\usepackage{authblk}
\author[1]{Anders Munch}
\author[1]{Thomas~A.~Gerds}
\affil[1]{Section of Biostatistics, University of Copenhagen}

@@

% \authormark{Munch et al.}

% \address[1]{\orgdiv{Section of Biostatistics}, \orgname{University of
%     Copenhagen}, \orgaddress{\street{Østerfarigmagsgade 5}, \postcode{2100}, \state{Copenhagen}, \country{Denmark}}}

% \corresp[$\ast$]{Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}

% \received{Date}{0}{Year}
% \revised{Date}{0}{Year}
% \accepted{Date}{0}{Year}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In right censored
  data, the choice of the loss function and the estimation of the expected loss
  need careful consideration. We introduce the state learner, a new super
  learner for survival analysis, which simultaneously evaluates libraries of
  prediction models for the event of interest and the censoring distribution.
  The state learner can be applied to all types of survival models, works in the
  presence of competing risks, and does not require a single pre-specified
  estimator of the conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its performance through
  numerical experiments. We illustrate the application of the state learner with
  prostate cancer data, as a stand-alone prediction tool, and, for causal
  inference, as a way to estimate the nuisance parameter models of a smooth
  statistical functional.
\end{abstract}

@@

We define the state learner, a new super learner for right-censored
data, which simultaneously estimates the expected loss of learners of
the event time distribution and the censoring distribution. The loss
function underlying the state learner operates on the coarsened
data. In the right censored survival setting the coarsened data
consist of the minimum and the order of the censoring time and the
event time as well as the baseline covariates. The state learner can
include a broad class of survival models as learners, it can handle
competing risks, and it does not require a single pre-specified
estimator of the conditional censoring distribution. To analyze the
theoretical properties of the state learner we focus on the discrete
super learner which combines the library of learners by picking the
one that minimises the cross-validated loss \citep{van2007super}. In
the presence of competing risks, our algorithm uses separate libraries
of learners for the cumulative hazard functions for each of the
competing risks and for the censoring distribution. We show that the
oracle selector of the state learner is consistent if all libraries
contain a consistent learner and prove a finite sample oracle
inequality.

@@

The result of the state learner algorithm could be a medical risk
prediction model \citep{gerds2021medical} which predicts the
probability of an event based on covariates in the presence of
competing risks. The other application is in targeted learning where
conditional event probabilities occur as high-dimensional nuisance
parameters which need to be estimated at a certain rate
\citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}.
For the asymptotic bias term of targeted estimator, which uses the
state learner to estimate nuisance parameters, we show that a second
order product structure holds. We illustrate both applications of the
state learner with prostate cancer data
\citep{kattan2000pretreatment}. We introduce our notation and
framework in Section~\ref{sec:framework}. In
Section~\ref{sec:super-learning} we define general super learning for
right-censored data. Section~\ref{sec:super-learner-simple} introduces
the state learner, and Section~\ref{sec:theor-results-prop} provides
theoretical guarantees. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted
learning. We report results of our numerical experiments in
Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
contains a discussion of the merits and limitations of our
proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.  An implementation of
the state learner is available at
\url{https://github.com/amnudn/statelearner} along with codes for
reproducing our numerical experiments.

@@

exclusive states. % (Figure \ref{fig:multi-state-process}).

@@

The asymptotic distribution of \( \hat{\Psi}^0 \) is
difficult to analyze due to the cross-validated model selection step
involved in the estimation of the nuisance parameters $\Lambda_1$ and
$\Lambda_2$.
% In addition, the estimator will typically have an
% asymptotic bias that vanishes at a too slow 
% rate.
Using tools from semi-parametric efficiency theory, it is possible to
construct a so-called targeted or debiased estimator with an
asymptotic distribution which we know how to estimate
\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
targeted estimator is based on the efficient influence function for
the parameter $\tilde{\Psi}^0$ and relies on estimators of the
nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
efficient influence function is a \( P \)-zero mean and square
integrable function which we denote by
\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
justified because any regular asymptotically linear estimator that has
\( \psi \) as its influence function is asymptotically efficient,
meaning that it has smallest asymptotic variance among all regular
asymptotically linear estimators \citep{bickel1993efficient}.

An example of a targeted estimator is the one-step estimator, defined as
\begin{equation}
    \label{eq:one-step-def}
    \hat{\Psi}_{\text{OS}}(\data_n)
    =
    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
    H_n(\data_n))
    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
      \hat{b}_n(\data_n) )]},
\end{equation}
where \( \empmeas \) is the empirical measure of a data set
\(\{O_i\}_{i=1}^n\).
Under suitable regularity conditions
we have the
following asymptotic expansion of the one-step estimator
\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric},
\begin{equation*}
  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
\end{equation*}
where the remainder term has the form
\begin{equation}
  \label{eq:4}
  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
  = \mathcal{O}_P{
    \left\{
      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
      +
      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
      +
      \|\Gamma-\hat{\Gamma}_{n}\|^2
    \right\}
  },
\end{equation}
for some suitable norm \( \|\blank \| \), for instance the
\( \mathcal{L}_{P}^2 \)-norm. When equation~(\ref{eq:4}) holds and the
nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
consistently estimated at rate \( \smallO_P{(n^{-1/4})} \), then
\begin{equation}
  \label{eq:3}
  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
\end{equation}
where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
calculating
\begin{equation*}
  \left[
    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
    \;
    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
  \right],
\end{equation*}
where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile of the standard normal
distribution and
\begin{equation*}
  \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
      \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
    \right]}.
\end{equation*}

@@

In this section we report results from a simulation study. In the first part we
consider estimation of the conditional survival function. Here we compare the
state learner to two IPCW based discrete super learners that use either the
Kaplan-Meier estimator or a Cox model to estimate the censoring probability
\citep{gonzalez2021stacked}. In the second part we compare the state learner to
the super learner proposed by \cite{westling2021inference}.

In both parts we use the same data-generating mechanism. We generate
data according to a distribution motivated from a real data set in
which censoring depends on the baseline covariates. We simulate data
based on the prostate cancer study of
\cite{kattan2000pretreatment}. The outcome of interest is the time to
tumor recurrence, and five baseline covariates are used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
(HT, yes/no) and clinical stage (CS, six values). The study was
designed such that a patient's radiation dose depended on when the
patient entered the study \citep{gerds2013estimating}. This in turn
implies that the time of censoring depends on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a
sensitivity analysis was conducted based on simulated data. Here we use the
same simulation setup, where event and censoring times are generated
according to parametric Cox-Weibull models estimated from the original
data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original
data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
simulation setting as ``dependent censoring''. We also considered a
simulation setting where data were generated in the same way, except
that censoring was generated completely independently. We refer to
this simulation setting as ``independent censoring''.

% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
For all super learners we use a library consisting of three learners:
The Kaplan-Meier estimator
\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
main effects \citep{cox1972regression, survival-package}, and a random
survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
same library to learn the outcome distribution and the censoring
distribution. Note that the three learners in our library of learners
can be used to learn the cumulative hazard functions of the 
outcome and the censoring distribution. This works by
training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
learner for the cumulative hazard function of the outcome to learn the
cumulative hazard function of the censoring time, we mean that the
learner is trained on \( \data_n^c \).

We compare the state learner to two IPCW based super learners: The
first super learner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
%
% | time | sim_setting | true_events | true_cens | at_risk |
% |------+-------------+-------------+-----------+---------|
% |   36 | original    |      24.619 |    61.853 |  25.774 |
% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
%
The IPCW super learners use the integrated Brier score up to a fixed time
horizon (36 months). The marginal risk of the event before this time horizon is
\(\approx 24.6\)\%. Under the ``dependent censoring'' setting the marginal
censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
``independent censoring'' setting the marginal censoring probability before this
time horizon is \( \approx 38.7 \)\%.

@@

\begin{figure}
  \centering %
  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figures~\ref{fig:zelefski-out}
and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
learner selected prediction models for both censoring and outcome which have
similar or higher IPA compared to the prediction models selected by
\texttt{survSL}.
\begin{figure}
  \centering %
  \includeFigCond[width=.8\linewidth]{experiment-fig-sl-survSL-out.pdf}
  \caption[]{For the risk prediction models of the outcome provided by each
    of the super learners, the IPA at the fixed time horizon is plotted against
    sample size. The results are averages across 1000 repetitions and the error
    bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-out}
\end{figure}

\begin{figure}
  \centering %
  \includeFigCond[width=.5\linewidth]{experiment-fig-sl-survSL-cens.pdf}
  \caption[]{For the risk prediction models of the censoring model provided by
    each of the super learners, the IPA at the fixed time horizon is plotted
    against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-cens}
\end{figure}

@@

% \begin{table}
  % \caption{\label{tab:zel-library}Overview of the five learners used by the
    % state learner when applied to the prostate cancer data set. The Kaplan-Meier
    % estimator was fitted using the package \texttt{prodlim}
    % \citep{Gerds_2019prodlim}. All Cox models included all five covariates in
    % the model and were fitted using the package \texttt{survival}
    % \citep{survival-package}. All penalised Cox models included all five
    % covariates as linear predictors and were fitted using the package
    % \texttt{glmnet} \citep{glmnet-cox,glmnet-glm}. The random forest was fitted
    % with the package \texttt{randomForestSRC} \citep{rfsrc-paclage}.}
  % \centering
  % \begin{tabular}{lll}
    % \toprule
    % Family & Model & Description \\
    % \midrule
    % Marginal & \texttt{KM} & The Kaplan-Meier estimator \\ 
    % Cox & \texttt{Cox} & All five covariates included with additive effects \\
           % & \( \texttt{Cox strata CS} \)  & Cox model stratified on CS  \\
    % Penalised Cox & \texttt{Lasso} & Cox model with \( L_1 \)-norm penalty   \\
           % & \texttt{Elastic} & Cox model with \( L_1 \)- and \( L_2 \)-norm penalty \\
    % Random forest & \texttt{RF} & Random forest with default settings \\
    % \bottomrule 
  % \end{tabular}
% \end{table}

>>>>

%% Packages
\usepackage{amsthm, dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}

@@

@@

\theoremstyle{thmstyleone}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

%% Line spacing
\linespread{1.65}

% \usepackage{lineno}
% \linenumbers

@@

\RequirePackage{graphicx}
\usepackage[font=small]{caption}

\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

% New operators and commands
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}

%% Title
\title{\Large The state learner -- a super learner for right-censored data}
\date{\small \today}

\usepackage{authblk}
\author[1,*]{\normalsize Anders Munch}
\author[1]{\normalsize Thomas~A.~Gerds}
\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}

\usepackage{fancyhdr}
\fancyhf{}
\fancyhead[l]{\textit{Munch et al.}}
\fancyhead[r]{\textit{The state learner}}
\fancyfoot[c]{\thepage}
\renewcommand\headrulewidth{0pt}
\pagestyle{fancy}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In right-censored
  data, the choice of the loss function and the estimation of the expected loss
  need careful consideration. We introduce the state learner, a new super
  learner for survival analysis, which simultaneously evaluates libraries of
  prediction models for the event of interest and the censoring distribution.
  The state learner can be applied to all types of survival models, works in the
  presence of competing risks, and does not require a single pre-specified
  estimator of the conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its performance through
  numerical experiments. We illustrate the application of the state learner with
  prostate cancer data, as a stand-alone prediction tool, and, for causal
  inference, as a way to estimate the nuisance parameter models of a smooth
  statistical functional.
\end{abstract}

@@

\textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
    estimation, right-censored data, super learner}}

\pagebreak

@@

We define the state learner, a new super learner for right-censored
data, which simultaneously estimates the expected loss of learners of
the event time distribution and the censoring distribution. The loss
function underlying the state learner operates on the coarsened
data. In the right-censored survival setting the coarsened data
consist of the minimum and the order of the censoring time and the
event time as well as the baseline covariates. The state learner can
include a broad class of survival models as learners, it can handle
competing risks, and it does not require a single pre-specified
estimator of the conditional censoring distribution. To analyse the
theoretical properties of the state learner we focus on the discrete
super learner which combines the library of learners by picking the
one that minimises the cross-validated loss \citep{van2007super}. In
the presence of competing risks, our algorithm uses separate libraries
of learners for the cumulative hazard functions for each of the
competing risks and for the censoring distribution. We show that the
oracle selector of the state learner is consistent if all libraries
contain a consistent learner and prove a finite sample oracle
inequality.

@@

The state learner algorithm can output a medical risk prediction model
\citep{gerds2021medical} which predicts the probability of an event based on
covariates in the presence of competing risks. The other application is in
targeted learning where conditional event probabilities occur as
high-dimensional nuisance parameters which need to be estimated at a certain
rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. For
the asymptotic bias term of targeted estimator, which uses the state learner to
estimate nuisance parameters, we show that a second order product structure
holds. We illustrate both applications of the state learner with prostate cancer
data \citep{kattan2000pretreatment}.

We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define
general super learning for right-censored data.
Section~\ref{sec:super-learner-simple} introduces the state learner, and
Section~\ref{sec:theor-results-prop} provides theoretical guarantees. In
Section~\ref{sec:targeted-learning} we discuss the use of the state learner in
the context of targeted learning. We report results of our numerical experiments
in Section~\ref{sec:numer-exper} and analyse a prostate cancer data set in
Section~\ref{sec:real-data-appl}. Section~\ref{sec:discussion} contains a
discussion of the merits and limitations of our proposal.
Appendices~\ref{sec:proof-proposition} and~\ref{sec:state-learner-with} contain
proofs. An implementation of the state learner is available at
\url{https://github.com/amnudn/statelearner} along with codes for reproducing
our numerical experiments.

@@

exclusive states: \( 0 \), \( 1 \), \( 2 \), or \( -1 \).
% (Figure \ref{fig:multi-state-process}).

@@

The asymptotic distribution of \( \hat{\Psi}^0 \) is
difficult to analyse due to the cross-validated model selection step
involved in the estimation of the nuisance parameters $\Lambda_1$ and
$\Lambda_2$.
% In addition, the estimator will typically have an
% asymptotic bias that vanishes at a too slow 
% rate.
Using tools from semi-parametric efficiency theory, it is possible to
construct a so-called targeted or debiased estimator with an
asymptotic distribution which we know how to estimate
\citep{bickel1993efficient,van2011targeted,chernozhukov2018double}. A
targeted estimator is based on the efficient influence function for
the parameter $\tilde{\Psi}^0$ and relies on estimators of the
nuisance parameters $\Gamma$, \( \Lambda_1 \) and $\Lambda_2$. The
efficient influence function is a \( P \)-zero mean and square
integrable function which we denote by
\( \psi(\blank ; \Lambda_1, \Lambda_2, \Gamma) \). The name is
justified because any regular asymptotically linear estimator that has
\( \psi \) as its influence function is asymptotically efficient,
meaning that it has smallest asymptotic variance among all regular
asymptotically linear estimators \citep{bickel1993efficient}.

An example of a targeted estimator is the one-step estimator, defined as
\begin{equation}
    \label{eq:one-step-def}
    \hat{\Psi}_{\text{OS}}(\data_n)
    =
    \tilde{\Psi}^0(\hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
    H_n(\data_n))
    + \empmeas{[\psi(\blank; \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n),
      \hat{b}_n(\data_n) )]},
\end{equation}
where \( \empmeas \) is the empirical measure of a data set
\(\{O_i\}_{i=1}^n\).
Under suitable regularity conditions
we have the
following asymptotic expansion of the one-step estimator
\citep{pfanzagl1985contributions,van2003unified,fisher2021visually,kennedy2022semiparametric},
\begin{equation*}
  \hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)
  =  \empmeas{[\psi(\blank ; \Lambda_1, \Lambda_2, \Gamma)]}
  +\mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)} + \smallO_{P}(n^{-1/2}),
\end{equation*}
where the remainder term has the form
\begin{equation}
  \label{eq:4}
  \mathrm{Rem}{(\hat{\Lambda}_{1,n},\hat{\Lambda}_{2,n},  \hat{\Gamma}_n, P)}
  = \bigO_P{
    \left\{
      \|\Lambda_1-\hat{\Lambda}_{1,n}\|^2
      +
      \|\Lambda_2-\hat{\Lambda}_{2,n}\|^2
      +
      \|\Gamma-\hat{\Gamma}_{n}\|^2
    \right\}
  },
\end{equation}
for some suitable norm \( \|\blank \| \), for instance the
\( \mathcal{L}_{P}^2 \)-norm. When equation~(\ref{eq:4}) holds and the
nuisance parameters $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
consistently estimated at rate \( \smallO_P{(n^{-1/4})} \), then
\begin{equation}
  \label{eq:3}
  \sqrt{n}(\hat{\Psi}_{\text{OS}}(\data_n)- \Psi(P)) \weakly \mathcal{N}(0,
  P{[\psi(\blank; \Lambda_1, \Lambda_2, \Gamma)^2]}),
\end{equation}
where we use \( \weakly \) to denote weak convergence \citep{van2000asymptotic}.
In particular, equation~(\ref{eq:3}) and Slutsky's lemma imply that we can
obtain asymptotically valid \((1-\alpha)\cdot100\%\) confidence intervals by
calculating
\begin{equation*}
  \left[
    \hat{\Psi}_{\text{OS}}(\data_n) - q_{\alpha/2} \hat{\sigma}(\data_n) ,
    \;
    \hat{\Psi}_{\text{OS}}(\data_n) + q_{\alpha/2} \hat{\sigma}(\data_n)
  \right],
\end{equation*}
where \( q_{\alpha} \) is the \( (1-\alpha) \)-quantile of the standard normal
distribution and
\begin{equation*}
  \hat{\sigma}(\data_n)^2 = \empmeas{ \left[ \psi(\blank;
      \hat{a}_{1,n}(\data_n), \hat{a}_{2,n}(\data_n), \hat{b}_n(\data_n))^2
    \right]}.
\end{equation*}

@@

In this section we report results from a simulation study where we consider
estimation of the conditional survival function. In the first part, we compare
the state learner to two IPCW based discrete super learners that use either the
Kaplan-Meier estimator or a Cox model to estimate the censoring probability
\citep{gonzalez2021stacked}. In the second part we compare the state learner to
the super learner proposed by \cite{westling2021inference}.

In both parts we use the same data-generating mechanism. We generate data
according to a distribution motivated from a real data set in which censoring
depends on the baseline covariates. We simulate data based on the prostate
cancer study of \cite{kattan2000pretreatment}. The outcome of interest is the
time to tumor recurrence, and five baseline covariates are used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum (GSS, values
between 6 and 10), radiation dose (RD), hormone therapy (HT, yes/no) and
clinical stage (CS, six values). The study was designed such that a patient's
radiation dose depended on when the patient entered the study
\citep{gerds2013estimating}. This in turn implies that the time of censoring
depends on the radiation dose. The data were re-analysed in
\citep{gerds2013estimating} where a sensitivity analysis was conducted based on
simulated data. Here we use the same simulation setup, where event and censoring
times are generated according to parametric Cox-Weibull models estimated from
the original data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original data
\citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this simulation
setting as `dependent censoring'. We also considered a simulation setting where
data were generated in the same way, except that censoring was generated
completely independently. We refer to this simulation setting as `independent
censoring'.

% ANDERS: hedder det ikke Nelson-Aalen i stedet for Kaplan-Meier?
%         kan du skrive klart hvad outcome er i denne foerste del? er det combined endpoint som man kalder for tumor-free survival?  
For all super learners we use a library consisting of three learners:
The Kaplan-Meier estimator
\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model with
main effects \citep{cox1972regression, survival-package}, and a random
survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the
same library to learn the outcome distribution and the censoring
distribution. Note that the three learners in our library of learners
can be used to learn the cumulative hazard functions of the 
outcome and the censoring distribution. The latter works by
training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a
learner for the cumulative hazard function of the outcome to learn the
cumulative hazard function of the censoring time, we mean that the
learner is trained on \( \data_n^c \).

We compare the state learner to two IPCW based super learners: The
first super learner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
%
% | time | sim_setting | true_events | true_cens | at_risk |
% |------+-------------+-------------+-----------+---------|
% |   36 | original    |      24.619 |    61.853 |  25.774 |
% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
%
The IPCW super learners use the integrated Brier score up to a fixed time
horizon (36 months). The marginal risk of the event before this time horizon is
\(\approx 24.6\)\%. Under the `dependent censoring' setting the marginal
censoring probability before the time horizon is \(\approx 61.9\)\%. Under the
`independent censoring' setting the marginal censoring probability before this
time horizon is \( \approx 38.7 \)\%.

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figures~\ref{fig:zelefski-out}
and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
learner selected prediction models for both censoring and outcome which have
similar or higher IPA compared to the prediction models selected by
\texttt{survSL}.
\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
  \caption[]{For the risk prediction models of the outcome provided by each
    of the super learners, the IPA at the fixed time horizon is plotted against
    sample size. The results are averages across 1000 repetitions and the error
    bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-out}
\end{figure}

\begin{figure}
  \centering %
  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
  \caption[]{For the risk prediction models of the censoring model provided by
    each of the super learners, the IPA at the fixed time horizon is plotted
    against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-cens}
\end{figure}

@@
